A comparison of design-based and model-based approaches for finite population spatial data.

### 2022-04-11 inst/manuscript/manuscript.Rmd
minor updates

---
title: A comparison of design-based and model-based approaches for finite population spatial data.
author:
  - name: Michael Dumelle
    affiliation: USEPA
    footnote: 1
  - name: Matt Higham
    affiliation: STLAW
  - name: Jay M. Ver Hoef
    affiliation: NOAA
  - name: Anthony R. Olsen
    affiliation: USEPA
  - name: Lisa Madsen
    affiliation: OSU
address:
  - code: USEPA
    address: United States Environmental Protection Agency, 200 SW 35th St, Corvallis, Oregon, 97333
  - code: STLAW
    address: Saint Lawrence University Department of Mathematics, Computer Science, and Statistics, 23 Romoda Drive, Canton, New York, 13617
  - code: NOAA
    address: Marine Mammal Laboratory, Alaska Fisheries Science Center, National Oceanic and Atmospheric Administration, Seattle, Washington, 98115
  - code: OSU
    address: Oregon State University Department of Statistics, 239 Weniger Hall, Corvallis, Oregon, 97331
footnote:
    code: 1
    text: "Corresponding Author: Michael Dumelle (Dumelle.Michael@epa.gov)"
abstract: |
  1.   The design-based and model-based approaches to frequentist statistical inference rest on fundamentally different foundations. In the design-based approach, inference relies on random sampling. In the model-based approach, inference relies on distributional assumptions. We compare the approaches for finite population spatial data.
  2. We provide relevant background for the design-based and model-based approaches and then study their performance using simulations and an analysis of real mercury concentration data. In the simulations, a variety of sample sizes, location layouts, dependence structures, and response types are considered. In the simualations and real data analysis, the population mean is the parameter of interest and performance is measured using statistics like bias, squared error, and interval coverage.
  3. When studying the simulations and mercury concentration data, we found that regardless of the strength of spatial dependence in the data, sampling plans that incorporate spatial locations (spatially balanced samples) generally outperform sampling plans that ignore spatial locations (non-spatially balanced samples). We also found that model-based approaches tend to outperform design-based approaches, even when the data are skewed (and by consequence, the model-based distributional assumptions violated). The performance gap between these approaches is small when spatially balanced samples are used but large when non-spatially balanced samples are used. This suggests that the sampling choice (whether to select a sample that is spatially balanced) is most important when performing design-based inference.
  4. There are many benefits and drawbacks to the design-based and model-based approaches for finite population spatial data that practitioners must consider when choosing between them. We provide relevant background contextualizing each approach and study their properties in a variety of scenarios, making recommendations for use based on the practitioner's goals.
journal: "Methods in Ecology and Evolution"
date: "`r Sys.Date()`"
bibliography: mybibfile.bib
linenumbers: true
numbersections: true
csl: elsevier-harvard.csl
preamble: >
  \usepackage{bm}
  \usepackage{bbm}
  \usepackage{color}
  \DeclareMathOperator{\var}{{var}}
  \DeclareMathOperator{\cov}{{cov}}
  \usepackage{caption}
  \usepackage{subcaption}
  \usepackage{setspace}
  \usepackage{fancyhdr}
  \pagestyle{fancy}
  \fancyhf{}
  \chead{Spatial design-based vs model-based}
  \doublespacing
output: rticles::elsevier_article
---

@@

The design-based approach assumes the population is fixed. Randomness is incorporated via the selection of population units according to a sampling design. A sampling design assigns a positive probability of inclusion (inclusion probability) in the sample to each population unit. These inclusion probabilities are later used to estimate population parameters. Some examples of commonly used sampling designs include simple random sampling, stratified random sampling, and cluster sampling.

When sampling designs incorporate spatial locations into sampling, we call the resulting samples "spatially balanced." One approach to selecting spatially balanced samples is the Generalized Random Tessellation Stratified (GRTS) algorithm [@stevens2004spatially], which we discuss in more detail in Section \ref{subsec:spb_design}. When sampling designs do not incorporate spatial locations into sampling, we call the resulting samples "non-spatially balanced."

@@

Instead of estimating fixed, unknown population parameters, as in the design-based approach, often the goal of model-based inference is to predict a realized variable, or value. For example, suppose the realized mean of all population units is the value of interest. Instead of a fixed, unknown mean, we are the value of the mean, a random variable. Prediction intervals are then derived using assumptions of the data-generating stochastic process. If we repeatedly generate response values from the same process and select samples, then 95% of all 95% prediction intervals constructed from a procedure with appropriate coverage will contain their respective realized means. @cressie1993statistics and @schabenberger2017statistical provide thorough reviews of model-based approaches for spatial data. In Fig. \ref{fig:fig1}, we provide a visual comparison of the design-based and model-based approaches (@verhoef2002sampling and @brus2021statistical provide similar figures).

@@

We previously mentioned that the design-based approach can be used to select spatially balanced samples (samples that incorporate spatial locations of the population units). Spatially balanced samples are useful because parameter estimates from these samples tend to vary less than parameter estimates from samples that are not spatially balanced [@stevens2004spatially; @barabesi2011sampling; @grafstrom2013well; @robertson2013bas; @wang2013design; @benedetti2017spatiallyreview]. The first spatially balanced sampling algorithm to see widespread use was the Generalized Random Tessellation Stratified (GRTS) algorithm [@stevens2004spatially]. To quantify the spatial balance of a sample, @stevens2004spatially proposed loss metrics based on Voronoi polygons (Dirichlet Tessellations). After the GRTS algorithm was developed, several other spatially balanced sampling algorithms emerged, including the Local Pivotal Method [@grafstrom2012spatially; @grafstrom2018spatially], Spatially Correlated Poisson Sampling [@grafstrom2012spatiallypoisson], Balanced Acceptance Sampling [@robertson2013bas], Within-Sample-Distance Sampling [@benedetti2017spatially], and Halton Iterative Partitioning Sampling [@robertson2018halton]. In this manuscript, we select spatially balanced samples using the Generalized Random Tessellation Stratified (GRTS) algorithm because it is readily available in the `spsurvey` \textbf{\textsf{R}} package [@dumelle2022spsurvey] and naturally accommodates finite and infinite sampling frames, unequal inclusion probabilities, and replacement units (replacement units are population units that can be sampled when a population unit originally selected can no longer be sampled).

@@

After determining $\mathcal{A}_k$, it is placed into hierarchical order. Hierarchical order is a numeric order that first sorts $\mathcal{A}_k$ by the level-one addresses from smallest to largest, then sorts $\mathcal{A}_k$ by the level-two addresses from smallest to largest, and so on. For example, $\mathcal{A}_2$ in hierarchical order is the set $\{00, 01, 02, 03, 10, ..., 13, 20, ..., 23, 30, ..., 33\}$. Because hierarchical ordering sorts by level-one cells, then level-two cells, and so on, population units that have similar hierarchical addresses tend to be nearby one another in space.Next each population unit is mapped to a one-dimensional line in hierarchical order where each population unit's inclusion probability equals its line-length. If a level-$k$ cell has multiple population units in it, they are randomly placed within the cell's respective line segment. A uniform random variable is then simulated in $[0, 1]$ and a systematic sample is selected on the line, yielding $n$ sample points for a sample size $n$. Each element in this systematic sample falls on some population unit's line segment, and thus that population unit is selected in the sample.

The GRTS algorithm selects samples by utilizing a particular mapping between two-dimensional and one-dimensional space that preserves proximity relationships. Via this mapping, units in two-dimensional space are partitioned using a hierarchical address. This hierarchical address is used to map population units to a one-dimensional line. On the one dimensional line, each population unit's line length equals its inclusion probability. Then, a systematic sample of population units is selected on the line and mapped back to two-dimensional space, yielding the desired sample. @stevens2004spatially provide more technical details.

@@

It is also important to quantify the uncertainty in $\hat{\tau}_{ht}$.  @horvitz1952generalization and @sen1953estimate provide variance estimators for $\hat{\tau}_{ht}$, but these estimators have two drawbacks. First, they rely on calculating $\pi_{ij}$, the probability that population unit $i$ and population unit $j$ are both in the sample -- this quantity can be challenging if not impossible to calculate analytically. Second, these estimators ignore the spatial locations of the population units. To address these two drawbacks simultaneously, @stevens2003variance proposed the local neighborhood variance estimator. The local neighborhood variance estimator does not rely on $\pi_{ij}$ and incorporates spatial locations -- for technical details see @stevens2003variance. @stevens2003variance show the local neighborhood variance estimator tends to reduce the estimated variance of $\hat{\tau}$ and yield more precise (narrower) confidence intervals compared to variance estimators that ignore spatial locations.

@@

Finite Population Block Kriging (FPBK) is a model-based approach that expands the geostatistical Kriging framework to the finite population setting [@verhoef2008spatial]. Instead of developing inference based on a specific sampling design, we assume the data are generated by a spatial stochastic process. We summarize some of the basic principles of FBPK next -- for technical details, see @verhoef2008spatial. Let ${\mathbf{z} \equiv \{\text{z}(s_1), \text{z}(s_2), . . . , \text{z}(s_N) \}}$ be an $N \times 1$ response vector at locations $s_1$, $s_2$, . . . , $s_N$ that can be measured at the $N$ population units. Suppose we want to use a sample to predict some linear function of the response variable, $f(\mathbf{z}) = \mathbf{b}^\prime \mathbf{z}$, where $\mathbf{b}^\prime$ is a $1 \times N$ vector of weights (e.g, the population mean is represented by a weights vector whose elements all equal $1 / N$). Denoting quantities that are part of the sampled population units with a subscript \emph{s} and quantities that are part of the unsampled population units with a subscript \emph{u}, let

@@

FBPK assumes $\bm{\delta}$ in Equation$~$\ref{equation:Zmarginal} has mean-zero and a spatial dependence structure that can be modeled using a covariance function. This covariance function is commonly assumed to be non-negative, second-order stationary (depending only on the distance between population units), isotropic (independent of direction), and decay with distance between population units [@cressie1993statistics]. Henceforth, it is implied that we have made these same assumptions regarding $\bm{\delta}$, though @chiles1999geostatistics, pp. 80-93 discuss covariance functions that are not second-order stationary, not isotropic, or not either. A variety of flexible covariance functions can be used to model $\bm{\delta}$ [@cressie1993statistics]; one example is the exponential covariance function (@cressie1993statistics provides a thorough list of spatial covariance functions). The $i,j$th element of the exponential covariance matrix, $\cov(\bm{\delta})$, is
\mbox{}
\begin{align}\label{equation:expcov}
\cov(\delta_i, \delta_j) = 
\begin{cases} 
\sigma^2_{1}\exp(-h_{i,j}/\phi) & h_{i,j} > 0 \\
\sigma^2_{1} + \sigma^2_2 & h_{i,j} = 0
\end{cases}
,
\end{align}
where $\sigma^2_{1}$ is the variance parameter quantifying the variability that is dependent (coarse-scale), $\sigma^2_{2}$ is the variance parameter quantifying the variability that is independent (fine-scale), $\phi$ is the range parameter measuring the distance-decay rate of the covariance, and $h_{i,j}$ is the Euclidean distance between population units $i$ and $j$. The proportion of variability attributable to dependent random error is $\sigma^2_{1} / (\sigma^2_{1} + \sigma^2_{2})$. Similarly, the proportion of variability attributable to independent random error is $\sigma^2_{2} / (\sigma^2_{1} + \sigma^2_{2})$. Finally we note that $\sigma^2_{1}$ and $\sigma^2_{2}$ are often called the partial sill and nugget, respectively.

With the above model formulation, the Best Linear Unbiased Predictor (BLUP) for $f(\mathbf{b}'\mathbf{z})$ and its prediction variance can be computed. While details of the derivation are in @verhoef2008spatial, we note here that the predictor and its variance are both moment-based, meaning that they do not rely on any distributional assumptions. Distributional assumptions are used, however, when constructing prediction intervals.

Other approaches, such as k-nearest-neighbors [@fix1989discriminatory; @ver2013comparison] and random forest [@breiman2001random], among others, could also be used to obtain predictions for a mean or total from finite population spatial data. Compared to the k-nearest-neighbors and random forest approach, we prefer FBPK because it is model-based and relies on theoretically-based variance estimators leveraging the model's spatial covariance structure, whereas k-nearest-neighbors and random forests use ad-hoc variance estimators [@ver2013comparison]. Additionally, @ver2013comparison studied compared FBPK, k-nearest-neighbors, and random forest in a variety of spatial data contexts, and FBPK tended to perform best.

@@

## Simulation Study {#sec:mm_sim}

We used a simulation study to investigate performance of four sampling-analysis combinations. The first sampling-analysis combination was IRS-Design. In IRS-Design, samples were selected with the Independent Random Sampling (IRS) algorithm. The IRS algorithm ignores the spatial locations of the population units, thus the IRS samples were not spatially balanced. In IRS-Design, samples were analyzed using the design-based approach via the Horvitz-Thompson mean estimator and an IRS variance estimator that ignored the spatial locations of the units in the sample. The second sampling-analysis combination was IRS-Model, where samples were selected with the IRS algorithm and analyzed using the model-based approach via Restricted Maximum Likelihood (REML) estimation [@patterson1971recovery; @harville1977maximum; @wolfinger1994computing]. The third sampling-analysis combination was GRTS-Design, where samples were selected with the GRTS algorithm and analyzed using the design-based approach via the Horvitz-Thompson mean estimator and the local neighborhood variance estimator (which does incorporate the spatial locations of the units in the sample). The fourth and final sampling-analysis combination was GRTS-Model, where samples were selected with the GRTS algorithm and analyzed using the model-based approach via REML estimation. These sampling-analysis combinations are also provided in Table \ref{tab:designanalysis}. Lastly we note that for both the IRS and GRTS samples, equal inclusion probabilities were assumed for all population units. When IRS assumes equal inclusion probabilities for all population units, the algorithm is equivalent to simple random sampling (SRS).

@@

@@

## Application {#sec:mm_app}

@@

## Simulation Study {#sec:r_sim}

@@

\begin{figure}
  \centering
  \includegraphics[width = 1\linewidth]{figures/mse_eff.jpeg}
  \caption{Relative MStdE in the simulation study for the four sampling-analysis combinations. The rows indicate the proportion of dependent error and the columns indicate the response type.}
  \label{fig:mse_eff}
\end{figure}

@@

## Application {#sec:r_app}

@@

There are several benefits and drawbacks of the design-based and model-based approaches for finite population spatial data. Some we have discussed, but others we have not, and they are worthy of consideration in future research.  Design-based approaches are often computationally efficient, while model-based approaches can be computationally burdensome, especially for likelihood-based estimation methods like REML that rely on inverting a covariance matrix. The design-based approach also more naturally handles binary data, free from the more complicated logistic regression framework commonly used to analyze binary data in a model-based approach. The model-based approach, however, can more naturally quantify the relationship between covariates (predictor variables) and the response variable. The model-based approach also yields estimated spatial covariance parameters, which help better understand the dependence structure in the stochastic process of study. Model selection is also possible using model-based approaches and criteria such as cross validation, likelihood ratio tests, or AIC [@akaike1974new]. Model-based approaches are capable of more efficient small-area estimation than design-based approaches by leveraging distributional assumptions in areas with few observed units. Model-based approaches can also compute unit-by-unit predictions at unobserved locations and use them to construct informative visualizations like smoothed maps. In short, when deciding whether the design-based or model-based approach is more appropriate to implement, the benefits and drawbacks of each approach should be considered alongside the particular goals of the study.

@@

>>>>

---
title: A comparison of design-based and model-based approaches for finite population spatial data.
author:
  - name: Michael Dumelle
    affiliation: USEPA
    footnote: 1
  - name: Matt Higham
    affiliation: STLAW
  - name: Jay M. Ver Hoef
    affiliation: NOAA
  - name: Anthony R. Olsen
    affiliation: USEPA
  - name: Lisa Madsen
    affiliation: OSU
address:
  - code: USEPA
    address: United States Environmental Protection Agency, 200 SW 35th St, Corvallis, Oregon, 97333
  - code: STLAW
    address: Saint Lawrence University Department of Mathematics, Computer Science, and Statistics, 23 Romoda Drive, Canton, New York, 13617
  - code: NOAA
    address: Marine Mammal Laboratory, Alaska Fisheries Science Center, National Oceanic and Atmospheric Administration, Seattle, Washington, 98115
  - code: OSU
    address: Oregon State University Department of Statistics, 239 Weniger Hall, Corvallis, Oregon, 97331
footnote:
    code: 1
    text: "Corresponding Author: Michael Dumelle (Dumelle.Michael@epa.gov)"
abstract: |
  1.   The design-based and model-based approaches to frequentist statistical inference rest on fundamentally different foundations. In the design-based approach, inference relies on random sampling. In the model-based approach, inference relies on distributional assumptions. We compare the approaches for finite population spatial data.
  2. We provide relevant background for the design-based and model-based approaches and then study their performance using simulated and real data. In the simulated data, a variety of sample sizes, location layouts, dependence structures, and response types are considered. In the simulated and real data, the population mean is the parameter of interest and performance is measured using statistics like bias, squared error, and interval coverage.
  3. When studying the simulated and real data, we found that regardless of the strength of spatial dependence in the data, the Generalized Random Tessellation Stratified (GRTS) algorithm, which explicitly incorporates spatial locations into sampling, tends to outperform the Simple Random Sampling (SRS) algorithm, which does not explicitly incorporate spatial locations into sampling. We also found that model-based approaches tend to outperform design-based approaches, even for skewed data where the model-based distributional assumptions are violated. The performance gap between these approaches is small GRTS samples are used but large when SRS samples are used. This suggests that the sampling choice (whether to use GRTS or SRS) is most important when performing design-based inference.
  4. There are many benefits and drawbacks to the design-based and model-based approaches for finite population spatial data that practitioners must consider when choosing between them. We provide relevant background contextualizing each approach and study their properties in a variety of scenarios, making recommendations for use based on the practitioner's goals.
journal: "Methods in Ecology and Evolution"
date: "`r Sys.Date()`"
bibliography: mybibfile.bib
linenumbers: true
numbersections: true
csl: elsevier-harvard.csl
preamble: >
  \usepackage{bm}
  \usepackage{bbm}
  \usepackage{color}
  \DeclareMathOperator{\var}{{var}}
  \DeclareMathOperator{\cov}{{cov}}
  \usepackage{caption}
  \usepackage{subcaption}
  \usepackage{setspace}
  \usepackage{fancyhdr}
  \pagestyle{fancy}
  \fancyhf{}
  \chead{Spatial design-based vs model-based}
  \doublespacing
output: rticles::elsevier_article
---

@@

The design-based approach assumes the population is fixed. Randomness is incorporated via the selection of population units according to a sampling design. A sampling design assigns a non-zero probability of inclusion (inclusion probability) in the sample to each population unit. These inclusion probabilities are later used to estimate population parameters. Some examples of commonly used sampling designs include simple random sampling, stratified random sampling, and cluster sampling.

When samples are chosen in a manner such that the layout of sampled units reflects the layout of the population units, we call the resulting sample "spatially balanced." By "reflecting the layout of the population units", we mean that if population units are concentrated in specific areas, the units in the sample should be concentrated in the same areas. Because spatially balanced samples reflect the layout of the population units, they are not necessarily "spread out" in space in some equidistant manner.

One approach to selecting spatially balanced samples is the Generalized Random Tessellation Stratified (GRTS) algorithm [@stevens2004spatially], which we discuss in more detail in Section \ref{subsec:spb_design}. When sampling designs do not incorporate spatial locations into sampling, we call the resulting samples "non-spatially balanced."

@@

Instead of estimating fixed, unknown population parameters, as in the design-based approach, often the goal of model-based inference is to predict a realized variable, or value. For example, suppose the realized mean of all population units is the value of interest. Instead of a fixed, unknown mean, we are predicting the value of the mean, a random variable. Prediction intervals are then derived using assumptions of the data-generating stochastic process. If we repeatedly generate response values from the same process and select samples, then 95% of all 95% prediction intervals constructed from a procedure with appropriate coverage will contain their respective realized means. @cressie1993statistics and @schabenberger2017statistical provide thorough reviews of model-based approaches for spatial data. In Fig. \ref{fig:fig1}, we provide a visual comparison of the design-based and model-based approaches (@verhoef2002sampling and @brus2021statistical provide similar figures).

@@

We previously mentioned that the design-based approach can be used to select spatially balanced samples. Spatially balanced samples are useful because parameter estimates from these samples tend to vary less than parameter estimates from samples that are not spatially balanced [@stevens2004spatially; @barabesi2011sampling; @grafstrom2013well; @robertson2013bas; @wang2013design; @benedetti2017spatiallyreview]. The first spatially balanced sampling algorithm to see widespread use was the Generalized Random Tessellation Stratified (GRTS) algorithm [@stevens2004spatially]. To quantify the spatial balance of a sample, @stevens2004spatially proposed loss metrics based on Voronoi polygons (Dirichlet Tessellations). After the GRTS algorithm was developed, several other spatially balanced sampling algorithms emerged, including the Local Pivotal Method [@grafstrom2012spatially; @grafstrom2018spatially], Spatially Correlated Poisson Sampling [@grafstrom2012spatiallypoisson], Balanced Acceptance Sampling [@robertson2013bas], Within-Sample-Distance Sampling [@benedetti2017spatially], and Halton Iterative Partitioning Sampling [@robertson2018halton]. In this manuscript, we select spatially balanced samples using the Generalized Random Tessellation Stratified (GRTS) algorithm because it is readily available in the `spsurvey` \textbf{\textsf{R}} package [@dumelle2022spsurvey] and naturally accommodates finite and infinite sampling frames, unequal inclusion probabilities, and replacement units (replacement units are population units that can be sampled when a population unit originally selected can no longer be sampled).

@@

After determining $\mathcal{A}_k$, it is placed into hierarchical order. Hierarchical order is a numeric order that first sorts $\mathcal{A}_k$ by the level-one addresses from smallest to largest, then sorts $\mathcal{A}_k$ by the level-two addresses from smallest to largest, and so on. For example, $\mathcal{A}_2$ in hierarchical order is the set $\{00, 01, 02, 03, 10, ..., 13, 20, ..., 23, 30, ..., 33\}$. Because hierarchical ordering sorts by level-one cells, then level-two cells, and so on, population units that have similar hierarchical addresses tend to be nearby one another in space.Next each population unit is mapped to a one-dimensional line in hierarchical order where each population unit's inclusion probability equals its line-length. If a level-$k$ cell has multiple population units in it, they are randomly placed within the cell's respective line segment. A uniform random variable is then simulated in $[0, 1]$ and a systematic sample is selected on the line, yielding $n$ sample points for a sample size $n$. Each element in this systematic sample falls on some population unit's line segment, and thus that population unit is selected in the sample. For further details regarding the GRTS algorithm, see @stevens2004spatially.

@@

It is also important to quantify the uncertainty in $\hat{\tau}_{ht}$.  @horvitz1952generalization and @sen1953estimate provide variance estimators for $\hat{\tau}_{ht}$, but these estimators have two drawbacks. First, they rely on calculating $\pi_{ij}$, the probability that population unit $i$ and population unit $j$ are both in the sample -- this quantity can be challenging if not impossible to calculate analytically for GRTS samples. Second, these estimators ignore the spatial locations of the population units. To address these two drawbacks simultaneously, @stevens2003variance proposed the local neighborhood variance estimator. The local neighborhood variance estimator does not rely on $\pi_{ij}$ and estimates the variance of $\hat{\tau}$ conditional on the random properties of the GRTS sample -- the idea being that this conditioning should yield a more precise estimate of $\hat{\tau}$. They show that the each observation's contribution to the overall variance is dominated by local variation. Thus the local neighborhood variance estimator is a weighted sum of variance estimates from each observation's local neighborhood. These local neighborhoods contain observation itself and its three nearest neighbors. For more details, see @stevens2003variance.

@@

Finite Population Block Kriging (FPBK) is a model-based approach that expands the geostatistical Kriging framework to the finite population setting [@verhoef2008spatial]. Instead of developing inference based on a specific sampling design, we assume the data are generated by a spatial stochastic process. We summarize some of the basic principles of FPBK next -- for technical details, see @verhoef2008spatial. Let ${\mathbf{z} \equiv \{\text{z}(s_1), \text{z}(s_2), . . . , \text{z}(s_N) \}}$ be an $N \times 1$ response vector at locations $s_1$, $s_2$, . . . , $s_N$ that can be measured at the $N$ population units. Suppose we want to use a sample to predict some linear function of the response variable, $f(\mathbf{z}) = \mathbf{b}^\prime \mathbf{z}$, where $\mathbf{b}^\prime$ is a $1 \times N$ vector of weights (e.g, the population mean is represented by a weights vector whose elements all equal $1 / N$). Denoting quantities that are part of the sampled population units with a subscript \emph{s} and quantities that are part of the unsampled population units with a subscript \emph{u}, let

@@

FPBK assumes $\bm{\delta}$ in Equation$~$\ref{equation:Zmarginal} has mean-zero and a spatial dependence structure that can be modeled using a covariance function. This covariance function is commonly assumed to be non-negative, second-order stationary (depending only on the distance between population units), isotropic (independent of direction), and decay with distance between population units [@cressie1993statistics]. Henceforth, it is implied that we have made these same assumptions regarding $\bm{\delta}$, though @chiles1999geostatistics, pp. 80-93 discuss covariance functions that are not second-order stationary, not isotropic, or not either. A variety of flexible covariance functions can be used to model $\bm{\delta}$ [@cressie1993statistics]; one example is the exponential covariance function (@cressie1993statistics provides a thorough list of spatial covariance functions). The $i,j$th element of the exponential covariance matrix, $\cov(\bm{\delta})$, is
\mbox{}
\begin{align}\label{equation:expcov}
\cov(\delta_i, \delta_j) = 
\begin{cases} 
\sigma^2_{1}\exp(-h_{i,j}/\phi) & h_{i,j} > 0 \\
\sigma^2_{1} + \sigma^2_2 & h_{i,j} = 0
\end{cases}
,
\end{align}
where $\sigma^2_{1}$ is the variance parameter that quantifies the spatially dependent variability, $\sigma^2_{2}$ is the variance parameter the quantifies that spatially independent variability, $\phi$ is the range parameter measuring the distance-decay rate of the covariance, and $h_{i,j}$ is the Euclidean distance between population units $i$ and $j$. In geostatistical literature, $\sigma^2_{1}$ is often called the partial sill and $\sigma^2_{2}$ is often called the nugget.

The parameters in Equation$~$\ref{equation:Zmarginal} can be estimated using a variety of techniques, but we focus on using restricted maximum likelihood [@patterson1971recovery; @harville1977maximum; @wolfinger1994computing]. REML is preferred over maximum likelihood (ML) because ML estimates can be badly biased for small sample sizes, due to the fact that ML makes no adjustment for the simultaneous estimation of $\bm{\beta}$ and $\bm{\delta}$ [@patterson1971recovery].  Minus twice the REML log-likelihood of the sampled sites is given by
\begin{equation}\label{equation:reml}
  \ln|\bm{\Sigma}| + (\bm{z}_s - \bm{X}_s \bm{\tilde{\beta}})^T \bm{\Sigma}_{ss}^{-1}(\bm{z}_s - \bm{X}_s \bm{\tilde{\beta}}) + \ln|\bm{X}_s^T \bm{\Sigma}_{ss}^{-1} \bm{X}_s| + (n - p) \ln(2 \pi) ,
\end{equation}
where $\bm{\tilde{\beta}} = (\bm{X}_s^T \bm{\Sigma}_{ss}^{-1} \bm{X}_s)^{-1} \bm{X}_s^T \bm{\Sigma}_{ss}^{-1} \bm{z}_s$ and $\bm{\Sigma}_{ss}$  is the covariance matrix of the sampled sites. Minimizing Equation$~$\ref{equation:reml} yields $\bm{\hat{\delta}}_{reml}$, the REML estimates of $\bm{\delta}$. Then $\bm{\beta}_{reml}$, the REML estimate of $\bm{\beta}$, is given by $(\bm{X}_s^T \bm{\hat{\Sigma}}_{ss}^{-1} \bm{X})^{-1} \bm{X}_s^T \bm{\hat{\Sigma}}_{ss}^{-1} \bm{z}_s$, where $\bm{\hat{\Sigma}}_{ss}$ is $\bm{\Sigma}_{ss}$ evaluated at $\bm{\hat{\delta}}_{reml}$.

With the model formulation in Equation$~$\ref{equation:Zmarginal}, the Best Linear Unbiased Predictor (BLUP) for $f(\mathbf{b}'\mathbf{z})$ and its prediction variance can be computed. While details of the derivation are in @verhoef2008spatial, we note here that the predictor and its variance are both moment-based, meaning that they do not rely on any distributional assumptions. Distributional assumptions are used, however, when constructing prediction intervals.

Other approaches, such as k-nearest-neighbors [@fix1989discriminatory; @ver2013comparison] and random forest [@breiman2001random], among others, could also be used to obtain predictions for a mean or total from finite population spatial data. Compared to the k-nearest-neighbors and random forest approach, we prefer FPBK because it is model-based and relies on theoretically-based variance estimators leveraging the model's spatial covariance structure, whereas k-nearest-neighbors and random forests use ad-hoc variance estimators [@ver2013comparison]. Additionally, @ver2013comparison compared FPBK, k-nearest-neighbors, and random forest in a variety of spatial data contexts, and FPBK tended to perform best.

@@

## Simulated Data {#sec:mm_sim}

We used a simulation study to investigate performance of four sampling-analysis combinations. The first sampling-analysis combination was IRS-Design. In IRS-Design, samples were selected with the Independent Random Sampling (IRS) algorithm. The IRS algorithm ignores the spatial locations of the population units, thus the IRS samples were not spatially balanced. In IRS-Design, samples were analyzed using the design-based approach via the Horvitz-Thompson mean estimator and an IRS variance estimator that ignored the spatial locations of the units in the sample. The second sampling-analysis combination was IRS-Model, where samples were selected with the IRS algorithm and analyzed using the model-based approach while estimating the covariance parameters ($\bm{\delta}$) and fixed effects ($\bm{\beta}$ using restricted maximum likelihood (REML). The third sampling-analysis combination was GRTS-Design, where samples were selected with the GRTS algorithm and analyzed using the design-based approach via the Horvitz-Thompson mean estimator and the local neighborhood variance estimator (which does incorporate the spatial locations of the units in the sample). The fourth and final sampling-analysis combination was GRTS-Model, where samples were selected with the GRTS algorithm and analyzed using the model-based approach while estimating the covariance parameters ($\bm{\delta}$) and fixed effects ($\bm{\beta}$ using restricted maximum likelihood (REML). These sampling-analysis combinations are also provided in Table \ref{tab:designanalysis}. Lastly we note that for both the IRS and GRTS samples, equal inclusion probabilities were assumed for all population units. When IRS assumes equal inclusion probabilities for all population units, the algorithm is equivalent to simple random sampling (SRS).

@@

CHANGE LOGNORMAL VERBAGE TO SKEWED

@@

\begin{figure}
\centering
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/symm_pop_hist.jpeg}
  \caption{}
  \label{fig:symm_pop_hist}
\end{subfigure}
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/skew_pop_hist.jpeg}
  \caption{}
  \label{fig:skew_pop_hist}
\end{subfigure}
\caption{Histograms of an example population realization for the normal (a) and skewed (b) scenarios for the simulated data.}
\label{fig:sim_pops}
\end{figure}

## National Lakes Assessment Data {#sec:mm_app}

@@

## Simulated Data {#sec:r_sim}

@@

@@

## National Lakes Assessment DAta {#sec:r_app}

@@

There are several benefits and drawbacks of the design-based and model-based approaches for finite population spatial data. Some we have discussed, but others we have not, and they are worthy of consideration in future research.  Design-based approaches are often computationally efficient, while model-based approaches can be computationally burdensome, especially for likelihood-based estimation methods like REML that rely on inverting a covariance matrix. The design-based approach easily handles binary data through a straightforward application of the Horvitz-Thompson estimator. In contrast, analyzing binary data using a model-based approach generally requires a logistic mixed regression model, which can be challenging to estimate and interpret [@bolker2009generalized]. The model-based approach, however, can more naturally quantify the relationship between covariates (predictor variables) and the response variable. The model-based approach also yields estimated spatial covariance parameters, which help better understand the dependence structure in the stochastic process of study. Model selection is also possible using model-based approaches and criteria such as cross validation, likelihood ratio tests, or AIC [@akaike1974new]. Model-based approaches are capable of more efficient small-area estimation than design-based approaches by leveraging distributional assumptions in areas with few observed units. Model-based approaches can also compute unit-by-unit predictions at unobserved locations and use them to construct informative visualizations like smoothed maps. In short, when deciding whether the design-based or model-based approach is more appropriate to implement, the benefits and drawbacks of each approach should be considered alongside the particular goals of the study.

@@

We would like to thank the editors and anonymous reviewers for their thoughtful comments which we feel greatly improved the manuscript.
